---
title: "LPA for Marc"
description: |
author:
  - name: Joseph Bulbulia
    url: https://josephbulbulia.netlify.app
    affiliation: Victoria University of Wellington
    affiliation_url: https://www.wgtn.ac.nz
    orcid_id: 0000-0002-5861-2056
date: 2021-MAR-23
output:
  distill::distill_article:
    self_contained: true
    toc: false
    code_folding: true
    highlight: tango
    highlight_downlit: false

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library("tidyverse") # data wrangling
library("tidyLPA") # latent profile analysis
library("patchwork") # making useful graphs 
library("table1") # tables
library("ggraph") # graphs
library("correlation")
library("report")
library("here") # file organisation
library("parameters") # data dimension reduction
library("DRR") # data dimension reduct
library("nFactors")
library("kableExtra")

```

## Data processing

Here, we perform a Latent Profile Analysis on a Marc's dataset 


First we import the data and rename the columns. 

```{r}
# read data. Because the data were prepared for MPLUS, I assume that there  are no column names, so we must create these. First read, and set "col_names" to FALSE

dat <-
  as_tibble(
    readr::read_csv(here::here("colleagues", "marc", "KDat2403.csv"), col_names = FALSE)
    )


# Next assign column names according to Marc's email dated March 23, 2021, at about 5pm  

df <- dat %>%
  dplyr::rename(
    "id" = "X1",
    "anx" = "X2",
    "cdct" = "X3",
    "sh" = "X4",
    "stht" = "X5",
    "satt" = "X6",
    "wb" = "X7",
    "rads" = "X8"
  )

# Vars are:
# PNum 
# Anx  - anxiety continuous
# Cdct - conduct problems continuous
# SH - self harm no or yes
# STht  - suicidal thought no or yes
# SAtt  - suicide attempt no or yes
# WB - wellbeing continuous
# radS - depression (RADS)
```


## PCA

We can reduce features of the data and *then* apply LPA. This is experimental `r knitr::asis_output("\U1F923")`



WARNING: I'm not sure it makes sense to perform an LPA on a PCA dataset, but here it goes

# Obtain PCA
```{r}
# dataset without id variable
ddf <- df %>%
  dplyr::select(-id)

# PCA
set.seed(12)
out <-
  parameters::reduce_parameters(ddf,
                                method = "PCA",
                                n = "max",
                                distance = "euclidean")

# Prepare the data 

# rename the columns
out1 <-  out %>%
  dplyr::rename(
    wb_neg_PCA =  "rads_-0.87/stht_-0.74/sh_-0.73/wb_0.72",
    satt_PCA = "satt_0.65",
    cdct_PCA = "cdct_-0.84",
    anx_PCA = "anx_0.66"
  )
```

### LPA

```{r cache = TRUE}
## fit an LPA on the Principle Components
fit4 <- out1 %>%
 dplyr::select(wb_neg_PCA,
         satt_PCA,
         cdct_PCA,
         anx_PCA) %>%
  tidyLPA::single_imputation() %>%  #handle missingness
  scale() %>%  # use standard deviation units
  tidyLPA::estimate_profiles(3:9)
```

### Results

```{r cache = TRUE, layout="l-body-outset",  fig.height=12, fig.width=12}
# Table of fits
tidyLPA::get_fit(fit4)
fit4 %>%
tidyLPA::compare_solutions(statistics = c("BIC", "Entropy"))
```

### Graph all fits

```{r  cache = T, layout="l-body-outset",  fig.height=8, fig.width=12}
tidyLPA::plot_profiles(fit4, add_line = TRUE) 
```

Inspecting the graph reveals many middling solutions. 

We might prefer a 4 or 5 class solution. 

Here is a 4 class solution

```{r  cache = T, layout="l-body-outset",  fig.height=8, fig.width=12}
tidyLPA::plot_profiles(fit4[[2]], add_line = TRUE) 
```

Calculate frequencies:


```{r cache = F}
d4 <- tidyLPA::get_data(fit4[[2]])
d4 %>%
  dplyr::select(Class) %>%
  dplyr::mutate(class = as.factor(Class)) %>%
  dplyr::group_by(class) %>%
  dplyr::summarise(n = n()) %>%
  dplyr::mutate(freq = n / sum(n)) %>%
  parameters::print_html() # html table
```
c
About 6 % are in the low wpPCA -> high sattPCA class. This is the same proportion as in the 3 class solution



### prcr


We can try a person centered analysis, see [prcr package webside](https://jrosen48.github.io/prcr/articles/introduction_to_prcr.html)


First we standardise our variables


```{r  cache = TRUE}
library(arm)
df2 <- df %>%
  dplyr::mutate(
    anx_s = as.numeric(scale(anx)),
    cdct_s = as.numeric(scale(cdct))  ,
    sh_s = as.numeric(scale(sh)),
    stht_s = as.numeric(scale(stht)),
    wb_s = as.numeric(scale(wb)),
    satt_s = as.numeric(scale(satt)),
    rads_s = as.numeric(scale(rads))
  )
```

prcr analysis and plot profiles
```{r}
library(prcr)
m1 <- create_profiles_cluster(df2, 
                              anx_s, 
                              cdct_s, 
                              sh_s, 
                              stht_s, 
                              wb_s,
                              satt_s,
                              rads_s,
                              n_profiles = 4)

plot_profiles(m1, to_center = T)
```



## Appendix 1 Results table for your publication 


```{r}
get_estimates(fit4[[2]])%>%
  print_html()
```

## Appendix 2  Material for your methods section


### Partial correlations of all variables
 
table
```{r  cache = TRUE, layout="l-body-outset",  fig.height=8, fig.width=12}
corr <- df %>%
  dplyr::select(-id) %>%
  correlation::correlation(partial = TRUE, multilevel = FALSE)


# Partial correlation table
corr %>%
  print_html()

# correlations graph
corr %>%
  plot()
```

###   Summary of variables

```{r layout="l-body-outset",  fig.height=8, fig.width=12}
skimr::skim(df, -id)
```

As evident from the histograms, there is there isn't much variation in the variables "sh", "stht", and "satt"

Let's look at the frequencies. 

```{r   cache = TRUE}
table1::table1(~
                 factor(sh) +
                 factor(stht) +
                 factor(satt),
               data = df
               )
```

6.4% of the sample are ticking 1 for suicide attempt. 





