---
title: "LPA for Marc"
description: |
author:
  - name: Joseph Bulbulia
    url: https://josephbulbulia.netlify.app
    affiliation: Victoria University of Wellington
    affiliation_url: https://www.wgtn.ac.nz
    orcid_id: 0000-0002-5861-2056
date: 2021-MAR-23
output:
  distill::distill_article:
    self_contained: true
    toc: false
    code_folding: true
    highlight: tango
    highlight_downlit: false

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library("tidyverse") # data wrangling
library("here") # file organisation
library("kableExtra") # tables
library("prcr") # analysis
library("sjlabelled") # remove labels
library("easystats") # tables //report

```

## Data processing

First we import the data and rename the columns. 

```{r}
# read data. Because the data were prepared for MPLUS, I assume that there  are no column names, so we must create these. First read, and set "col_names" to FALSE

dat <-
  as_tibble(
    readr::read_csv(here::here("colleagues", "marc", "KDat2403.csv"), col_names = FALSE)
    )


# Next assign column names according to Marc's email dated March 23, 2021, at about 5pm  

df <- dat %>%
  dplyr::rename(
    "id" = "X1",
    "anx" = "X2",
    "cdct" = "X3",
    "sh" = "X4",
    "stht" = "X5",
    "satt" = "X6",
    "wb" = "X7",
    "rads" = "X8"
  )
# remove lables
df <- sjlabelled::remove_all_labels(df)

# Vars are:
# PNum 
# Anx  - anxiety continuous
# Cdct - conduct problems continuous
# SH - self harm no or yes
# STht  - suicidal thought no or yes
# SAtt  - suicide attempt no or yes
# WB - wellbeing continuous
# radS - depression (RADS)


# standardise variables, so that differences are in standard deviation units

df2 <- df %>%
  dplyr::mutate(
    anx_s = as.numeric(scale(anx)),
    cdct_s = as.numeric(scale(cdct))  ,
    sh_s = as.numeric(scale(sh)),
    stht_s = as.numeric(scale(stht)),
    wb_s = as.numeric(scale(wb)),
    satt_s = as.numeric(scale(satt)),
    rads_s = as.numeric(scale(rads))
  )
```

## Person Centred Analysis 
For details about Person centered analysis see: 

See: https://jrosen48.github.io/prcr/


```{r}
# Vignette
# Provides an easy-to-use yet adaptable set of tools to conduct person-center analysis using a two-step clustering procedure. As described in Bergman and El-Khouri (1999) <doi:10.1002/(SICI)1521-4036(199910)41:6%3C753::AID-BIMJ753%3E3.0.CO;2-K>, hierarchical clustering is performed to determine the initial partition for the subsequent k-means clustering procedure.

library(prcr)

m1 <- create_profiles_cluster(df2, 
                              anx_s, 
                              cdct_s, 
                              sh_s, 
                              stht_s, 
                              wb_s,
                              satt_s,
                              rads_s,
                              n_profiles = 4)
print.prcr <- function(x){
    cat("$clustered_processed_data\n\n")
    print(x$clustered_processed_data)
} 

```


## Results of a four cluster analaysis

Using person-centered clustering analysis, we find that 6% of the sample can be classified into a suicide-attempt cluster (cluster 3).  The cluster that is prone to suicidal attempts is 1.5 standard deviations higher in both suicidal rumination and self-harm. Members of this cluster tend to be almost 1 standard deviation higher than average in conduct problem as well as depression. Members of this class are reliably lower in well-being when compared with the sample average. 

We also detect another cluster that over 1.5 times higher in self-harm but does not attempt suicide. This cluster represents 45% of the sample. Members of this cluster are prone to suicidal rumination, and present lower in wellbeing; they present higher in depression, anxiety, and misconduct. 

To infer suicidal attempt hazard among youth across New Zealand would require a combination of time-series data and stratified sample weighting. Nevertheless, this analysis reveals that both suicidal idea and self-harm is prominent among those who report attempting suicide and among those who do not, and that suicidal ideal and tendencies to self harm are correlated. 

```{r layout="l-body-outset", fig.width=12, fig.height=10}
prcr::plot_profiles(m1) + theme_minimal() + labs(title = "PC")
```


```{r}
# Frequency of classes
m1%>%
  dplyr::select(cluster)%>%
  dplyr::mutate(cluster = as.factor(cluster)) %>%
  dplyr::group_by(cluster)%>%
  dplyr::summarise(n = n())%>%
  dplyr::mutate(freq = n / sum(n)) %>%
  parameters::print_html() # html table
  
```



## Appendix 1 

We can write a predictive model for reported suicide attempts (yes/no)


```{r}
sjPlot::tab_model(
  m2<-glm(sh ~ anx_s + cdct_s +  stht_s + wb_s + rads_s, data = df2, family = binomial))
```

We do not observe substantial multiple collinearity in the predictors:

```{r}
performance::check_collinearity(m2)
```





<!-- ## Summary -->

<!-- ```{r} -->
<!-- summary(m1) %>% -->
<!--   kbl(caption = "Results") %>% -->
<!--   kable_classic_2(c("striped", "hover"), full_width = TRUE)  -->
<!-- ``` -->




