---
title: "Notes_Daryl_Meaning"
author: "Joseph Bulbulia"
date: "12/02/2021"
output: html_document
---


```{r global_options, include = FALSE}
knitr::opts_chunk$set(
  fig.path = "figs/",
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  echo = FALSE,
  fig.width = 10,
  fig.height = 8
)
#results="hide",
#fig.width= 8,
#fig.height=,
#fig.fullwidth = TRUE)
```

```{r load_libraries, include=FALSE}
library("tidyverse")
library("dplyr")
library("easystats")
library("lme4")
library("splines")
library("mgcv")
library("table1")
library("ggplot2")
library("ggeffects")
library("gghighlight")
library("Amelia")
library("magrittr")
library("patchwork")
library("sjPlot")
library("texreg")
library("tinytex")
library("kableExtra")
library("ggpubr")
library("insight")
#library("viridis")
#library("ggpubr")
#library("wesanderson")
library("ggsci")
library("report")
library("papaja")
library("here")
library("citr")
library("here")
library("brms")
library("rstan")
library("rstanarm")
library("tidybayes")
library("dagitty")
library("ggdag")
#library("furniture") # messes with table1
library("ggdag")
library("ggplot2")
library("lubridate")
library("viridisLite")
library("bayesplot")
library("msm")
library("broom")
#library("magick")
# rstan options
rstan_options(auto_write=TRUE)
options(mc.cores=parallel::detectCores ())
#theme_set(theme_few())
```

```{r loaddata, include =FALSE, cache =TRUE}
#load data
df <- readRDS("~/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/00Bulbulia\ Pubs/2021/DATA/ldf.5")
```


We are interested in understanding the relationship between Meaning of Life (meaning) and Kessler 6 distress (distress) and also of Religious Identification (religious id) and Kessler 6 distress.

## Causal Model for distress ~ meaning. 

We begin by writing a causal model that we hope to evaluation for the relationship between meaning and distress. (See chapter 5-6 of [Statistical Rethinking](https://github.com/rmcelreath/stat_rethinking_2020))

In this model:
 
 - distress is predicted by religion + meaning + income + heath + social belonging + male
 - religion is predicted by age + male
 - meaning is predicted by religion + employed + male + age
 - income is predicted by employed
 - health is predicted by income and social belonging
 - social beloning is predicted by religion + male 

More complicated models are possible, and others might be more realistic (see the note at the end.) However we'll start here:


```{r, echo = TRUE}
library(ggdag)
dg<-dagify(d ~ r + m + i + h + sb + male + cv, 
       r ~ a + male,
       m ~ r + e + male + a,
       i ~ e + cv,
       e ~ cv,
       h ~ i + sb,
       sb ~ r + male + cv, 
       labels = c(
    "d" = "well being",
    "r" = "religion",
    "m" = "meaning",
    "i" = "income",
    "h" = "health",
    "sb" = "social belonging",
    "cv" = "covid timeline",
    "e" = "employed",
    "a" = "age",
    "male" = "male"),
    exposure = "m",
    outcome = "d")


ggdag::ggdag_paths(dg, text = FALSE, use_labels = "label")+ theme_dag_blank()

```

Note that there are many collider confounding relationships in the assumed causeal network

```{r, echo = TRUE}
ggdag_collider(dg, text = FALSE,use_labels = "label") + theme_dag_blank()
```

However, we can obtain an an unbiased estimate for the relationship between meaning and distress by including employed, religion, and "male" or income, male and religion and covid timeline. We'll choose employed because it is measured with less error (however income would be more precise... which adjustment set to choose will depend on our interests, here the effects are so large it won't matter... )

```{r, echo = TRUE}
ggdag_adjustment_set(
  dg,
  text = FALSE,
  use_labels = "label",
  shadow = TRUE,
  exposure =  "m",
  outcome =   "d",
) + theme_dag_blank()

```

Let's run the model 
```{r, show = FALSE, cache = TRUE}
# data
ord_dates_class <- c("Early_March_2020", "BaseLine_2018_19_w10", "Pre_COVID_2019_w11", "Jan_Feb_2020", "L4_Lockdown_2020","Post_Lockdown_2020")
cv <- df %>%
  dplyr::filter(YearMeasured == 1) %>%
  dplyr::filter(Wave == 2018 | Wave == 2019) %>%
  dplyr::group_by(Id) %>% filter(n() > 1) %>%
  dplyr::filter(n() != 0) %>%
  dplyr::ungroup(Id) %>%
  dplyr::mutate(Covid_Timeline =
                 # ifelse(
                 # TSCORE %in% 3837:3894,
                 # "Baseling",
                    as.factor(ifelse(
                      TSCORE %in% 3895:3921,
                      "Early_March_2020",
                      ifelse(
                        TSCORE %in% 3921:3954,
                        "L4_Lockdown_2020",
                        ifelse(
                          TSCORE > 3954,
                          "Post_Lockdown_2020",
                          ifelse(
                            TSCORE %in% 3842:3894,
                            "Jan_Feb_2020",
                          ifelse(
                            TSCORE %in% 3665:3842 &
                              Wave == 2019,
                            "Pre_COVID_2019_w11",
                            "BaseLine_2018_19_w10"
                          )
                        )
                      )
                    ) 
                 ) ))%>%
  dplyr::mutate(PrePostCovid =
                  ifelse(
                    TSCORE > 3895, "post_covid",
                    ifelse(
                      TSCORE < 3895 &
                        Wave == 2019,
                      "before_COVID_wave11",
                      "baseline_wave10"
                    )
                  ))%>% 
  dplyr::select(
    TSCORE,
    Wave,
    PrePostCovid,
    years,
    Id,
    Hours.Work,
    Household.INC,
    Hours.Exercise,
    LIFEMEANING,
    KESSLER6,
    #  K6sum,
    KESSLER6sum,
    Age,
    Male,
    Relid,
    Pol.Orient,
    Urban,
    GenCohort,
    Edu,
    Employed,
    YearMeasured,
    Covid_Timeline,
    HLTH.SleepHours,
    HLTH.Fatigue,
    LIFESAT,
    PWI,
    EthnicCats,
    NZdep,
    Partner,
    Pol.Orient,
    Urban,
    Euro,
    Religious,
    NZdep,
    WSCORE,
    Religion.Church,
    Emp.JobSecure,
    Alcohol.Intensity,
    Alcohol.Frequency
  ) %>%
  dplyr::group_by(Id) %>% filter(n() > 1) %>%
  dplyr::filter(n() != 0) %>%
  dplyr::ungroup(Id) %>%
  dplyr::filter(!is.na(KESSLER6sum),
                !is.na(LIFEMEANING),
                !is.na(Male),
                !is.na(Relid)) %>%
  dplyr::mutate(
    Urban = (Urban),
    Euro = (Euro),
    Religious = as.factor(Religious),
    Covid_Timeline = as.factor(Covid_Timeline)
  ) %>%
  dplyr::filter(!is.na(Relid),
                !is.na(LIFEMEANING))%>%
  dplyr::mutate(
    WSCORE = as.factor(WSCORE),
    EduC = scale(as.numeric(Edu), scale = F, center = TRUE),
    AgeDecadeC = scale(Age, scale = F, center = T) / 10,
    HoursWork10_C = scale(Hours.Work, scale = F, center = T) / 10,
    HoursSleep_C = scale(HLTH.SleepHours, scale=FALSE, center=TRUE),
    HoursExercise_C = scale(Hours.Exercise, scale=FALSE, center=TRUE),
    K6sum = as.integer(KESSLER6 * 6),
    k_S = scale(K6sum),
    LIFEMEANING_S = scale(LIFEMEANING),
    NZdepS = scale(NZdep),
    PolOrientS = scale(Pol.Orient, scale = T, center = T),
    Relid_S = scale(Relid, scale = T, center = T),
    Conservative = Pol.Orient,
    Conservative_S = PolOrientS,
    Religion_Church_S = scale(Religion.Church),
    Emp.JobSecure_S = scale(Emp.JobSecure),
    KESSLER6sum_S = scale(KESSLER6sum),
    KESSLER6sum = as.integer(KESSLER6sum),
    Alcohol.Intensity_S = scale(Alcohol.Intensity),
    Alcohol.Frequency_S = scale( Alcohol.Frequency),
    Hours.Work = as.integer(Hours.Work),
    Hours.Exercise = as.integer(Hours.Exercise),
    Household.INC_log_C = scale(log(Household.INC+1), center = TRUE, scale = FALSE)
  )%>%
  dplyr::mutate(Covid_Timeline = forcats::fct_relevel(Covid_Timeline, ord_dates_class))

#table(cv$Covid_Timeline)
# 
# 

cv1 <- cbind(cv,
             demean(
               cv,
               select = c(
                 "Emp.JobSecure",
                 "KESSLER6sum",
                 "HLTH.SleepHours",
                 "HLTH.Fatigue",
                 "Religion.Church",
                 "Hours.Work",
                 "Hours.Exercise",
                 "Relid",
                 "Religion.Church",
                 "LIFEMEANING"
               ),
               group = "Id"
             ))
```


Meaning on K6 distress
```{r cache = TRUE}
dm1<- lmer(k_S ~ bs(LIFEMEANING_S) + bs(Relid_S) + Employed + Male + (1|Id), data = cv1)
```

Here are the results
```{r cache = TRUE}
sjPlot::tab_model(dm1)
```

This graph ,however, is more informative.  We find a *massive* effect of meaning of life on distress. Plotting all the points, we can see that small increments of increasing in Meaning of Life predict large reductions in K6.  At the low end tail, the low meaning of life is almost synonmous with distress. Yet the questions are very different. (JB to say more... )

```{r cache = TRUE}
plot(ggeffects::ggpredict(dm1, terms = "LIFEMEANING_S [all]"), ci=.89,add.data = TRUE, dot.alpha = 0.01, jitter = .2 ) 
```

##  Causal model for distress ~ religious identification

Next, we use a multi-level non-linear model to estimate the relationship between religious identification and distress.  The DAG is the same as for distress, so the colliders with be the same as above, however this time we are interested in religious id as the exposure.



```{r}
# Same DAG
d1<-dagify(d ~ r + m + i + h + sb + male + cv, 
       r ~ a + male,
       m ~ r + e + male + a,
       i ~ e + cv,
       e ~ cv,
       h ~ i + sb,
       sb ~ r + male + cv, 
       labels = c(
    "d" = "well being",
    "r" = "religion",
    "m" = "meaning",
    "i" = "income",
    "h" = "health",
    "sb" = "social belonging",
    "cv" = "covid timeline",
    "e" = "employed",
    "a" = "age",
    "male" = "male"),
    exposure = "r",
    outcome = "d")
# same colliders
#ggdag_collider(dg1, text = FALSE,use_labels = "label") + theme_dag_blank()

# only condition on male for estimating the effect of Religion on d
ggdag_adjustment_set(
  d1,
  text = FALSE,
  use_labels = "label",
  shadow = TRUE,
  exposure =  "r",
  outcome =   "d",
) + theme_dag_blank()
```
The adjustment set says that for an unbiased estimate, we should condition on **age** and  **male.**

## Aside

Out of curiosity, what would happen if we "controlled" for meaning of life in this model?  The answer is that we activate collider biases.  We would then see an association open between religion and male, and we could then not obrtain a biased estimate of religion on distress. Let's return to this below.

```{r}
# ggdag_adjustment_set(
#   dg1,
#   text = FALSE,
#   use_labels = "label",
#   shadow = TRUE,
#   exposure =  "r",
#   outcome =   "d",
# ) + theme_dag_blank()
ggdag_dseparated(d1, controlling_for = "m", "a", "male") + theme_dag_blank()

```

This example illustrates the importance of expressing and explicit causal model before assessing a causal theory.  



## Back to distress ~ religious identification  
Let's return to assessing the relationship between religious identification and Kessler 6 distress.  We opt for a non-linear model b high religious identification results in huge differences not expressed at the low end of the religious ID scale. 


```{r cache = TRUE}
dm2<- lmer(k_S ~ bs(Relid_S) + AgeDecadeC  + Male + (1|Id),  data = cv1)
```

We find some sort of effect. However, non-linear coefficients in tables are opaque.

```{r}
sjPlot::tab_model(dm2)
```

Therefore, to understand what is happening, we need to graphing the results:

```{r  cache = TRUE}
plot(ggeffects::ggpredict(dm2, terms = "Relid_S [all]"), ci=.89, add.data = TRUE, dot.alpha = 0.01, jitter = .2 ) 
```

We do not find that religion is strongly associated with Kessler 6 distress. However, there is some evidence for slight easing of distress at the high end of the religious identification scale. Given noise in the data, we shouldn't make too much of this. 



## Model for meaning ~ religious identification

Writing the dag, we can obtain an unbiased estimate of religion on meaning by including **age** and **male**.

```{r}
#What do we condition on with religion
dg2<-dagify(m ~ r + e + male + a,
       r ~ a + male,
       e ~ ct, 
       labels = c(
    "r" = "religion",
    "m" = "meaning",
    "sb" = "social belonging",
    "a" = "age",
    "e" = "employed",
    "ct" = "Covid Timeline",
    "male" = "male"),
    exposure = "r",
    outcome = "m")

```

```{r}
# only condition on male for estimating the effect of Religion on d
ggdag_adjustment_set(
  dg2,
  text = FALSE,
  use_labels = "label",
  shadow = TRUE,
  exposure =  "r",
  outcome =   "m",
) + theme_dag_blank()

```

Running that model we observe that high religious identification predicts high levels of life_meaning. 

```{r cache=TRUE}
dm3<- lmer(LIFEMEANING_S ~ bs(Relid_S) + AgeDecadeC + Male + (1|Id),  data = cv1)
```

The coefficient table:

```{r}
sjPlot::tab_model(dm3)
```

And the predicted effects:

```{r}
plot(ggeffects::ggpredict(dm3, terms = "Relid_S [all]"), ci=.89, add.data = TRUE, dot.alpha = 0.01, jitter = .2 )
```


So we find that religion strongly predicts life meaning at the high end of the religion spectrum. 


## Ilusion that Religion Causes Distress 

Let's return to the first model above in which we estimated the effects of meaning of life adjusting for religion, but this time, let's graph the expected effects of religion within this model.  Recall we shouldn't do this because to model the association of religion on distress while including male + meaning+ employed invites collider bias.

Graphing both models

```{r cache = TRUE}
# model without meaning 
m2 <-plot(ggeffects::ggpredict(dm2, terms = c("Relid_S [all]")), ci=.89, add.data = TRUE, dot.alpha = 0.01, jitter = .2 ) 

# model with meaning
m1<-plot(ggeffects::ggpredict(dm1, terms = c("Relid_S [all]")), ci=.89, add.data = TRUE, dot.alpha = 0.01, jitter = .2 ) 

m2 + m1 + plot_annotation(title = "Comparision of Predicted Effects of Religion", subtitle = "a: model without meaning; b: mdoel with meaning", tag_levels = 'a') + plot_layout(guides = 'collect')
```


Note that religious identification is associate with "higher" distress in the model that include meaning. Why? 

This result is a reminder for how important it is to understand what a regression model is doing, and a reminder that causal effects cannot be read from regression tables. In the model in which we are including both meaning and religious identification, we are focusing on religious identification at different levels of meaning. The predicted responses is setting meaning to the population average. The b graph in focusing on the expected relationship of religion to distress among the population that scores average in levels of meaning.  However, we know that this is an unusual population because most people who are high on religion also tend to be high on meaning of life.  Moreover, by including meaning into a model that tests the causal assocation of religion with kessler 6, we introduce confounding. 

In any case, if we examine the predicted effects of Religion on Kessler 6 at different levels of Life Meaning, we find that meaning of life is driving the effects.


```{r cache = TRUE}
plot(ggeffects::ggpredict(dm1, terms = c( "Relid_S [all]", "LIFEMEANING_S[all]")), ci=.89, add.data = TRUE, dot.alpha = 0.01, jitter = .2 ) 
```



## Short summary
We can infer the following: 

- to the extend that religious identification increases a sense of meaning,  this is expected to diminish psychological distress. 
- We find high levels of meaning at the high end of religious identification. 
- increasing religious identification by itself is not a recipe for reducing psychological distress. 
- we do not really show that meaning of life is a cause of higher or lower psychological distress. We test the predictions of such a model, but it is possible that psychological distress affects perceptions of life's meaning. For example, if one were depressed...
- To test the causal direction model  we can look to natural experiments in which people who differ in meaning of life but are otherwise similar encounter distressing life events remain stable or change in response to distressing life events. If people with higher meaning of life show greater buffering, this would indicate that meaning protects people from distress.  Time series data in the NZAVS will allow us to do that (eg unemployment during COVID...)...  TBA


## Multilevel mediation model

We can combine the models and separate out their effects in a multilevel mediation analysis. To be done when time permits. 

```{r, echo = TRUE}
library(brms)
dat<-cv1%>%
  dplyr::mutate(mS = LIFEMEANING_S,
         rS = Relid_S,
         r = as.integer(Relid))

# I am not sure what montonic grouping effects would be, even, so wil use standardised indicators. 

path_y <- bf(
 k_S ~ mo(r) + mS + Employed + Male + ( rS + mS |i| Id ))

path_m <- bf(
 mS ~ mo(r) + AgeDecadeC + Male + ( rS |i| Id ))


fit1 <- brm(
  path_m + path_y + set_rescor(FALSE),
  data = dat,
  chains = 4,
  cores = 12,
 # control = list(adapt_delta = .99),
  file = here("fit1")
)

```



## A final thought on causal modelling

A slight adjustment to our causal assumptions would leave a model for the expected effects of meaning of life on distress **unidentifiable**! For example, if we were to imagine that there wereunmeasured community effects on both religion and meaning a sense of meaning, as follows:

```{r, echo = TRUE}
dg4<-dagify(d ~ r + m + i + h + sb + male + ct, 
       r ~ a + male + U,
       m ~ r + e + male + U,
       i ~ e,
       e ~ ct, 
       sb ~ r + male, 
       h ~ i + sb,
       labels = c(
    "d" = "well being",
    "r" = "religion",
    "m" = "meaning",
    "i" = "income",
    "h" = "health",
    "sb" = "social belonging",
    "a" = "age",
    "ct" = "Covid Timeline",
    "male" = "male",
    "U" = "Unmeasured Community Effects"),
    exposure = "m",
    outcome = "d",
    latent = "U")
# same colliders
ggdag_collider(dg4, text = FALSE,use_labels = "label") + theme_dag_blank()
```

Then we we would find the path between meaning and distress could not be closed with these indicators. 


```{r}
ggdag_adjustment_set(
  dg4,
  text = FALSE,
  use_labels = "label",
  shadow = TRUE,
  exposure =  "r",
  outcome =   "d",
) + theme_dag_blank()
```



It is worrying that we might expect such community effects both to exist, and to play important roles.  We'll need to approach inference with care.

## FOLLOW UP ON MEDIATION

These re the paths by which meaning might causally effect well-being on the model above



```{r, echo = TRUE}
library(ggdag)
dg3<-dagify(d ~ r + i + h + sb + male + cv + m, 
       r ~ a + male,
       m ~ r + e + male + a,
       i ~ e + cv,
       e ~ cv,
       h ~ i + sb,
       sb ~ r + male + cv, 
       labels = c(
    "d" = "well being",
    "r" = "religion",
    "m" = "meaning",
    "i" = "income",
    "h" = "health",
    "sb" = "social belonging",
    "cv" = "covid timeline",
    "e" = "employed",
    "a" = "age",
    "male" = "male"),
    exposure = "m",
    outcome = "d")


ggdag::ggdag_paths(dg3, from = "m", to = "d", text = FALSE, use_labels = "label")+ theme_dag_blank()

ggdag::ggdag_status(dg3,use_labels = "label")+ theme_dag_blank()
ggdag::ggdag_exogenous(dg3,use_labels = "label")+ theme_dag_blank()

```







<!-- ## Aside -->

<!-- Though if we make a slight tweak, and consider that meaning and community might reciprocally effect each other: -->

<!-- ```{r} -->
<!-- dg5<-dagify(d ~ r + m + i + h + sb + male + ct,  -->
<!--        r ~ a + male, -->
<!--        m ~ r + e + male, -->
<!--        U ~~ m, -->
<!--        U ~ r, -->
<!--        i ~ e, -->
<!--        e ~ ct,  -->
<!--        sb ~ r + male,  -->
<!--        h ~ i + sb, -->
<!--        labels = c( -->
<!--     "d" = "well being", -->
<!--     "r" = "religion", -->
<!--     "m" = "meaning", -->
<!--     "i" = "income", -->
<!--     "h" = "health", -->
<!--     "sb" = "social belonging", -->
<!--     "a" = "age", -->
<!--     "ct" = "Covid Timeline", -->
<!--     "male" = "male", -->
<!--     "U" = "Unmeasured Community Effects"), -->
<!--     exposure = "m", -->
<!--     outcome = "d", -->
<!--     latent = "U") -->

<!-- #  colliders -->
<!-- ggdag_collider(dg5, text = FALSE,use_labels = "label") + theme_dag_blank() -->
<!-- ``` -->

<!-- Then we would only need to condition on "male".  -->

<!-- ```{r} -->
<!-- ggdag_adjustment_set( -->
<!--   dg5, -->
<!--   text = FALSE, -->
<!--   use_labels = "label", -->
<!--   shadow = TRUE, -->
<!--   exposure =  "r", -->
<!--   outcome =   "d", -->
<!-- ) + theme_dag_blank() -->
<!-- ``` -->

<!-- To me this sort of fragility in our modeling approach -- in any modelling approach -- signals a need for genuine intellectual humility about what our data can reveal -- at least by present lights.  -->
