---
title: "How to avoid counfounding in Hedwig's Study? It depends on which DAG we assume"
author: "Joseph Bulbulia"
date: "12/01/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(#fig.path="figs/", 
                      message=FALSE, 
                      warning=FALSE,
                      collapse =TRUE,
                      echo=TRUE, 
                      #results="hide", 
                      fig.width= 10,
                      fig.height=8,
                      tidy = "styler")
```
                      
## Question

Does Psychopathy affect JobLevel over time? 


## Which variables do we need to control for?

We assume that 

  - Job levels are affected by Years + Psychopathy + Age + Male + Education + Time in Job + "U" (an unmeasured effect like priviledge)
  
  - Pyschopathy is affected by Male
  
  - JobTime is affected by Years + Psychopathy ("impulsivit") + Unmeasured Privilege
  
  - Education is affected by Time (more time to be educated) + Unmeasured Privilege. 
  
  - Age is affeced by Years
  
  - Male doesn't change and U doesn't change.
  
  - The Year effect stratifies on different levels of Psychopathy (i.e. Psychopathy won't be invariate over the years)


Now let's write out our causal model with these assumptions, and see which variables we need to include to understand the relationship between Psychopathic Personality and JobLevel. 




```{r}
library("dagitty")
library("ggdag")
dag0<- dagify(
  JobLevel ~ Years + Psycho ,
  JobLevel ~ Age + Male + Edu + JobTime  + U,
  Psycho ~ Male + Years,
  JobTime ~ Years + Psycho + U,
  Edu ~ Years + U,
  Age ~ Years,
  exposure =  "Psycho",
  latent = "U",
  outcome =   "JobLevel",
  labels = c(
    "Psycho" = "Psychopathy",
    "JobLevel" = "JobLevel",
    "Male" = "Male",
    "Edu" = "Edu",
    "JobTime" = "JobTime",
    "Years" = "Years",
    "Age" = "Age",
    "U" = "Unmeasured_Priviledge"
  )
) 

ggdag_adjustment_set(
  dag0,
  text = FALSE,
  use_labels = "label",
  shadow = TRUE,
  exposure =  "Psycho",
  outcome =   "JobLevel",
) + theme_dag_blank()

ggdag::ggdag_collider(dag0)
ggdag::ggdag_collider(dag0)
```



```{r}
# aaron use
library("ggdag")
ar<- dagify(
  jobsat ~ age + edu + actorP + partnerP + jobT ,
  actorP ~  male,
  partnerP ~ male,
  op ~ age + edu + actorP + partnerP + jobT,
  jobT ~ actorP + partnerP + unemploy,
  unemploy ~ actorP + partnerP,
  exposure =  "partnerP",
  outcome =   "op",
  labels = c(
    "jobsat" = "jobsatisfaction",
    "op" = "occupational prestige",
    "edu"  = "education",
    "unemploy" = "unemployment",
    "actorP" = "actor psychopathy",
    "partnerP" = "partner psychopathy",
    "jobT" = "job tenure",
     "male" = "male",
    "age" = "age"
  )
) 
dag_paths(ar)
ggdag_adjustment_set(
  ar,
  text = FALSE,
  use_labels = "label",
  shadow = TRUE,
  exposure =  "partnerP",
  outcome =   "op"
) + theme_dag_blank() + theme_dag_blank()

ggdag::ggdag_collider(ar,
                      stylized = TRUE,
                      text = FALSE,
                      use_labels = "label") + theme_dag_blank()


ggdag_dseparated(ar,
    from = "partnerP",
    to = "op",
    controlling_for = c( "male"),
    collider_lines = TRUE,
    text = FALSE,
    use_labels = "label"
  ) + theme_dag_blank()

```

```{r}

# AARONS
library("dagitty")
library("ggdag")

ag0<- dagify(
  js_f ~ time + p_f + p_o + jt_f + male ,
  p_f ~ male,
  p_o ~ male,
  jt_f ~ time,
  exposure =  "p_o",
 # latent = c("U_s","U_o"),
  outcome =   "js_f",
  labels = c(
    "js_f" = "Prof Success Focal",
   # "js_o" = "Prof Sucess Other",
    "p_f" = "psychopathy_focal",
    "p_o" = "psychopathy_other",
   "time" = "time",
    "jt_f" = "Job Time Focal",
     "male" = "male"
  )
) 

ggdag_adjustment_set(
  ag0,
  text = FALSE,
  use_labels = "label",
  shadow = TRUE,
  exposure =  "p_o",
  outcome =   "js_f",
) + theme_dag_blank()

ggdag::ggdag_collider(ag0)


ggdag_dseparated(ag0,
    from = "p_o",
    to = "js",
    controlling_for = c( "p_s",  "male"),
    collider_lines = TRUE,
    text = FALSE,
        use_labels = "label"
  ) + theme_dag_blank()


```


The minimal set is "Male" and "Years." 

What would happen if we were to roll with the "casual salad" approach and include the other variables that might affect Job Level? 

```{r}
ggdag_dseparated(dag0,
    from = "Psycho",
    to = "JobLevel",
    controlling_for = c( "Years",  "Male",  "Edu" , "JobTime", "U", "Age"),
    text = FALSE,
        use_labels = "label"
  ) + theme_dag_blank()

```


The answer is that we open collider confounds, at least on the assumptions of this DAG.  Of course, another DAG might be better, and you should feel free to play with this. 


## Our question might differ in which case we'd need a different model
Note, if we wanted to predict Job Level accurately, we *would* want the causal salad model. 


## What if we think that all the control variables might affect how psychopathy affects job performance?

Bad luck! No way to close the backdoors unless we can find another instrumental variable. 

```{r}
dag03 <- dagify(
  JobLevel ~ Years + Psycho + Age + Male + Edu + JobTime  + U,
  Psycho ~ Years  +  Age + Male + Edu + JobTime  + U,
  JobTime ~ Years + Psycho + U,
  Edu ~ Years + U,
  Age ~ Years,
  labels = c(
    "Psycho" = "Psychopathy",
    "JobLevel" = "JobLevel",
    "Male" = "Male",
    "Edu" = "Edu",
    "JobTime" = "JobTime",
    "Years" = "Years",
    "Age" = "Age",
    "U" = "Unmeasured_Priviledge"
  )
)

ggdag_adjustment_set(
  dag03,
  text = FALSE,
  use_labels = "label",
  shadow = TRUE,
  exposure =  "Psycho",
  outcome =   "JobLevel",
) + theme_dag_blank()

```

I doubt we believe such a model, though it might be implied by our decision to include all "control" covariates. In any case, we need to think about causation when adding "controls."


## Round 2 

```{r}
ar<-dagify(
  ps1 ~ py1 + py2 + reldur + jobdur + male, 
  py1 ~~ py2,
  py1 ~ male,
  py2 ~ male,
  reldur ~ py1 + py2,
  jobdur ~ py2 + py1, 
  exposure = "py2",
  outcome = "ps1"
)

ggdag_collider(ar)+ theme_dag_blank()
ggdag::ggdag_adjustment_set(ar, node_size = 14) +
  theme(legend.position = "bottom") + theme_dag_blank()
ggdag_drelationship(ar,collider_lines = TRUE, controlling_for = c("py1","py2","male","jobdur","reldur")) + theme_dag_blank()


```


<!-- Below is a more complicated version of the same.  -->

<!-- ## Below is ignorable. -->

<!-- We can write more complicated graphs ... but the ideal is the same: just to need to include Male.  -->

<!-- ## Which variables do we need to "control" for confounds? -->


<!-- Our interest is in causation not predictive accuracy.   -->

<!-- For this reason we must take care when including control ariables that obtain unbiasted estimate for the effect of psychopathic personality on promotion. -->

<!-- First we need to find the adjustment set for assessing whether psychopathy predicts promotion.  -->

<!-- ```{r} -->
<!-- if (!require(dagitty)) { -->
<!--   install.packages( "dagitty" ) -->
<!-- } -->

<!-- if (!require(ggdag)) { -->
<!--   install.packages( "ggdag" ) -->
<!-- } -->

<!-- library("dagitty") -->
<!-- library("ggdag") -->
<!-- dag2<-dagify( -->
<!--   Promotion ~  Success0 +  JOBTIME1 + Edu1 + Psycho +  MALE + U, -->
<!--   Success0 ~ JOBTIME0 + Edu0 + MALE + Psycho, -->
<!--   JOBTIME1 ~ JOBTIME0 + Edu1 + Psycho, # impulsivity?  -->
<!--   JOBTIME0 ~  Edu0 + Psycho, -->
<!--   Psycho ~ MALE, # unmeasured variable afffects both success and education  -->
<!--   Success0 ~ U,# unmeasured variable afffects both success and education, call it "priviledge" -->
<!--   Edu0 ~ U, -->
<!--   Edu1 ~ U, -->
<!--   Edu1 ~ Edu0, -->
<!--     labels = c( -->
<!--     "Promotion" = "Promotion", -->
<!--     "Success0" = "Success0", -->
<!--     "JOBTIME1" = "JOBTIME1", -->
<!--     "Edu1" = "Edu1", -->
<!--     "Psycho" = "Psycho", -->
<!--     "MALE" = "MALE", -->
<!--      "U" = "Unmeasured", -->
<!--     "JOBTIME0" = "JOBTIME0", -->
<!--     "Edu0" = "Edu0" -->
<!--     ) -->
<!--   )  -->
<!-- ``` -->


<!-- Plot: -->

<!-- ```{r} -->
<!-- ggdag_adjustment_set(dag2, -->
<!--     text = FALSE, -->
<!--     use_labels = "label", -->
<!--     shadow = TRUE, -->
<!--     exposure =  "Psycho", -->
<!--     outcome =   "Promotion", -->
<!--   ) + theme_dag_blank() -->

<!-- ``` -->

<!-- We observe that to assess the question of which variables to include to assess the effect of pyschopathic personality on promotion we need to condition on Male.  Note that we don not need to include more than Male to assess whether psychopathy predicts promotion, according to this causal DAG.  -->

<!-- We might ask, would including the other variables introduce confounding? After all, we might also want better prediction. The answer is yes. Including the other measures opens confounding from the unmeasured source of success.  -->


<!-- Plot: -->

<!-- ```{r} -->
<!-- ggdag_dseparated(dag2, -->
<!--     from = "Psycho", -->
<!--     to = "Promotion", -->
<!--     controlling_for = c( "Success0" , "U", "MALE", "Edu1","Edu0","JOBTIME0","JOBTIME1"),  -->
<!--     text = FALSE, -->
<!--     use_labels = "label", -->
<!--     collider_lines = TRUE, -->
<!--   ) + theme_dag_blank() -->
<!-- ``` -->


<!-- What if we inlcude only MALE, as our DAG instructed us? The collider confounding closes.  -->

<!-- ```{r} -->
<!-- ggdag_dseparated(dag2, -->
<!--     from = "Psycho", -->
<!--     to = "Promotion", -->
<!--     controlling_for = c( "MALE"),  -->
<!--     text = FALSE, -->
<!--     collider_lines = TRUE, -->
<!--     use_labels = "label") + theme_dag_blank() -->

<!-- ``` -->

<!-- What if we were to include only a subset of "control" variables?  Perhaps adjusting for level of success at baseline is important. Perhaps people at the bottom of the heap will not be promoted as quickly as those at the top, especially if they are psychopaths.  -->


<!-- ```{r} -->
<!-- ggdag_dseparated(dag2, -->
<!--     from = "Psycho", -->
<!--     to = "Promotion", -->
<!--     controlling_for = c( "MALE", "Success0"),  -->
<!--     text = FALSE, -->
<!--     collider_lines = TRUE, -->
<!--     use_labels = "label") + theme_dag_blank() -->

<!-- ``` -->


<!-- We need to close the door.  If we include only MALE we do that. -->

<!-- ```{r} -->
<!-- ggdag_dseparated(dag2, -->
<!--     from = "Psycho", -->
<!--     to = "Promotion", -->
<!--     controlling_for = c( "MALE"),  -->
<!--     text = FALSE, -->
<!--     collider_lines = TRUE, -->
<!--     use_labels = "label") + theme_dag_blank() -->

<!-- ``` -->


<!-- ## Summmary -->
<!-- As of today, my sense is that to understand whether Psychopathy causes promotion, we would be wise to consider only conditioning only on MALE. Doing so closes the backdoor between unmeasured influences on promotion that might also affect previous success that operate differently for those who identify as male and those who do not.  -->

<!-- It is a worry that small changes to a DAG will make big differences to which causal feaures of the worlk we think our results are describing.  -->

<!-- I don't know how close the DAGS I've presented here are to reality. I use them to illustrate the need to proceed with caution when introducing control variables.  -->


<!-- NOTE: -->
<!-- Causal inference gets much more complicated in a heirarchical modelling setting such as ours. -->

<!-- https://bpspsychub.onlinelibrary.wiley.com/doi/abs/10.1111/bmsp.12217 -->



