---
title: "Comments on `Internalising and Externalising in Parents and Offspring: The Impact of Sex and Early Life Adversity` by  Emerald O’Neill-Murchison"
description: |
author:
  - name: Joseph Bulbulia
    url: https://josephbulbulia.netlify.app
    affiliation: Victoria University of Wellington
    affiliation_url: https://www.wgtn.ac.nz
    orcid_id: 0000-0002-5861-2056
date: June 7, 2021
output:
  distill::distill_article:
    self_contained: true
    toc: false
    code_folding: true
    highlight: tango
    highlight_downlit: false
bibliography: references.bib
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
# set up
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = TRUE
)
```

```{r message = FALSE, include = FALSE}
# required packages
if (!require(ggdag)) install.packages("ggdag")

library("brms")
library("rstan")
library("rstanarm")
library("tidybayes")
library("bayesplot")
library("easystats")
library("tidyverse")
# rstan options
rstan_options(auto_write=TRUE)
options(mc.cores=parallel::detectCores ())
theme_set(theme_classic())
```

## Overview

### Aims

Forensic psychologists use the term "externalising" to denote
psychological disorders characterised by disinhibition or aggression.

Forensic psychologists use the term "internalising" to denote cognitive
and affective disorders.

Developmental forensic psychologists have investigated the etiology by
which specific behavioural and cognitive/affective disorders arise
during childhood, how offspring are affected by parents, and whether
there are sex differences.

Emerald's thesis extends previous investigations examining the
interrelated effects of parental pathology, early life adversity and sex
differences on adult offspring internalising and externalising. Her
thesis is distinctive in three ways.

First, Emerald's thesis "takes focus on the overall dimensions of
internalising and externalising as opposed to specific disorders"
(p.18). That is, the project approaches psychological disorders using
the conceptual frameworks of internalisation and externalisation.

Second, Emerald's thesis examines the relationships between parental
internalising and externalising, early childhood adversity (ELA), and
sex differences in the expression of adult offspring internalising and
externalising. Previous studies have not focussed on the interdependence
of these parameters within the same model.

Third, this project examines sex differences: specifically whether
paternal or maternal internalising or externalising differently affects
male and female offspring in adulthood.

### Importance

The study is important for the reasons that Emerald states:

> Young people's exposure and experiences of parental externalising, as
> well as ELA, can result in serious negative and lasting consequences
> for individuals and society. It is important to gain a clearer
> understanding of these impacts. As such, the aim of this project is to
> gain further understanding of the relationship between parental
> externalising and internalising and externalising as mediated by ELA;
> that is investigating whether ELA statistically accounts for the
> relationship between our predictor variable, parental externalising,
> and our dependent variables, internalising and externalising in
> offspring (p.18).

### Hypotheses

The project proposes three hypotheses:

1.  " ELA [will] positively account for the relationship between
    parental externalising and internalising and externalising" (p.19).
2.  "$\dots$ there [will] be no [statistically] significant difference
    overall in the impact of parental externalising between mothers and
    fathers, wherein both mothers and fathers alike significantly
    [affect] both male and female offspring with their externalising
    behaviour" (p.19)
3.  (a.) Maternal internalising [will have] a relatively greater effect
    on female offspring compared with male offspring (in adulthood) in
    both externalising and internalising and (b) Paternal externalising
    [will have] a relatively greater effect on male offspring compared
    with female offspring (in adulthood).

The motivation for these hypotheses is grounded in previous research.

### Findings

Emerald reports the following results:

1.  No direct association was demonstrated between maternal
    externalising or paternal externalising and internalising or
    externalising in offspring.
2.  ELA was associated with maternal externalising and paternal
    externalising in females only.
3.  The relationship between paternal externalising and internalising in
    male offspring was mediated by ELA, as was the relationship between
    maternal externalising and internalising in male offspring. The
    relationship between parental externalising and internalising in
    females however was not mediated by ELA.

## Evaluation

### Evaluation of the interpretation of the results.

1.  **No direct association was demonstrated between maternal
    externalising or paternal externalising and internalising or
    externalising in offspring.** I do not share this interpretation. I
    note that the result comes from a mediation model that also
    conditions on Early Life Adversity. As such, once we include ELA,
    the parental effects drop out. This result is what we would expect
    if causal effect of parental pathology is fully mediated by
    childhood experiences. So far as I can tell, the conclusions of the
    thesis are simply misstated. Before submitting the final version, I
    would ask that Emerald and her supervisors review the results. To
    help clarify my point, I have simulated data in Appendix 1.
2.  **The direct internalising and externalising predicts ELA in females
    only.** The first thing to notice here is that main effects and and
    interactions of sex on ELA were not tested within the same model.
    Instead, for reasons that remain unclear, Emerald appears to have
    performed a separate analyses on females and males, which strongly
    compromises our ability to infer sex differences. Second, the
    finding that parental pathology is unrelated to ELA is a strong
    indication of problems in the measurement approach. Having parents
    who are checked out or abusive would itself appear to be a
    dimensions of Early Life Adversity. Before submitting the final
    version, I would ask that Emerald conduct an analysis in only one
    model that includes male and female (See further comments below).
3.  **The relationship between paternal externalising and internalising
    in male offspring was mediated by ELA, as was the relationship
    between maternal externalising and internalising in male offspring.
    The relationship between parental externalising and internalising in
    females however was not mediated by ELA**. This finding is
    compromised in three ways. First, ELA is reliably related to
    offspring outcomes except in Group 2 Males. Second, as far as I can
    tell parental effects are fully mediated by ELA (see below) -- or at
    least, a correct interpretation is lacking. Third, as mentioned, the
    subgroup analyses of males and females relies on an incorrectly
    specified model.

### Clarity of purpose

I found the study challenging to follow. Definitions of the key terms,
externalisation and internalisation, do not occur until page 8. These
terms should have been defined in the abstract and again on the first
page of the introduction .

There were many acronyms throughout - CD, ACE, ASPD, PTSD, GAD, MDD,
OCD. The acronyms introduce heavy cognitive burdens, and many proved to
be side-points to the main interest: Early Life Adversity (ELA). The
thesis should be re-written without these acronyms. Indeed I would
encourage Emerald to write out "Early Life Adversity."

Before submitting a final draft, I would ask Emerald and her supervisors
to work on the clarify of the thesis's verbal expression and
readability.

### Context for research

The research topic is important, and Emerald makes a good argument for
its importance.

### Appropriate methods

Worryingly, the key inference is, as far as I can make out, incorrect.
Emerald writes: "We hypothesised ELA would positively account for the
relationship between parental externalising and internalising and
externalising." However, if the hypothesis were true, then including ELA
in the model would result in the conditional independence of parental
externalising/internalising and offspring externalising/internalising
(after adjusting for sex differences.) So we would expect that the
direct effect of paternal influence would drop out, \*\*as observed.\*\*
To show this, see my simulation in Appendix 1.

As such, I do not believe the following result is warranted:

> "Contrary to past research which has demonstrated many times that
> parental externalising confers a higher risk of offspring developing
> externalising behaviour, in our study we did not find a significant
> relationship between either maternal or paternal externalising and
> externalising symptoms in male or female offspring (Besemer, 2014;
> Foley et al., 2001; Furtado et al., 2006; Herndon & Iacono, 2005; Kim
> et al., 2009; Long et al., 2018; Whitten et al., 2019). This is in
> direct opposition to the strong links that have been drawn between
> parental externalising and the development of substance use problems
> in offspring, as well as behaviour difficulties and/or the engagement
> of offending behaviour (Besemer et al., 2017; Long et al., 2018).
> Furthermore, though there is an abundance of studies which report a
> strong association between ELA and externalising symptoms across the
> life course, our results did not replicate this finding in females or
> in males (see review by Curran et al., 2018). p.47

It should have been a tip off that the present study was incongruous
with past research, and I will confess to being disappointed that
Emerald did not systematically address her curious results. The stakes
in this research are, after all, important.

Also, as mentioned, the selection of two random groups and the running
of two parallel analyses on the two groups is a misunderstanding of
Pohlmann (2004). The approach is confusing, and unnecessarily reduces
the studies power. I think Emerald was hoping to cross-validate here
results, but this requires a different approach.

Also, as mentioned, the main effects of sex and interactions of sex and
the mediator (ELA) as well as sex and parental exposure were not
included in the same model. Instead Emerald performed a subgroup
analysis:

> Additionally, as we were interested in the sex-of-participant effect;
> whether males and females were impacted by maternal and paternal
> externalising, and ELAs in similar or different ways, analysis was
> **run by group**. This allowed us to compare the results of males and
> females. p.19

I could not find any justification for the sub-group analyses. To infer
differences Emerald needs to include sex-differences within the same
statistical model. And indeed on the final page, Emerald notes this (!),
"It must be noted that this conclusion has been drawn on the basis of
comparing the direct and indirect effects of the SEM model rather than a
specific test of moderation." p.49.

Presently, all inferences about differences among men and women are
invalid. I would ask Emerald and her supervisors to revisit these
inferences before Emerald lodges her thesis with the Victoria Library.

### Successful outcomes

The thesis, in its present form, does not appear to be successful.
However, I am sanguine: without too much trouble, the thesis can be
fixed.

### Effective presentation

The literature review is solid. The topic is well-motivated. However, I
struggled with assessing this thesis because I found myself regularly
confused by it's logic and clarity of expression. For this reason I do
not judge the thesis in its present form to be particularly effective.
However Emerald can work on its clarity of expression (see my "specific
points to consider" below for some directions).

## How to fix the several faults with the current manuscript

There's much to be happy about in this thesis. The literature review is
informative. Writing out the causal assumptions of the presumed model I
find that the variables that were selected to assess the causal
contributions of parental disorder on child disorder were correct (See
Appendix 2). Good work.

I believe that the thesis can be fixed with relatively little effort
should Emerald takes the following steps:

1.  Define the key theoretical terms "externalising" and "internalising"
    in the abstract and in the opening section of the introduction.
2.  Offer a clear theoretical justification for why the lumping strategy
    of psychological disorders into the large categories of
    "externalising" and "internalising" is motivated. Presently a clear
    rational is lacking.
3.  Do not adopt the sample splitting strategy wrongly attributed to
    Pohlmann (2004). While splitting would be OK for the measurement
    models, Pohlmann's advice does not apply to the inferential model.
    The splitting of the data set is confusing: it does not offer the
    cross-validated model that Emerald hopes for.
4.  Do not split the samples further into females and males. To
    investigate non-linear differences include both the main effects and
    interactions of sex. Note: in my view, stratifying by sex would be
    sufficient to assess the causal effects of parental disorders on
    offspring disorders. While I do not believe a "moderated mediation"
    analysis is motivated for causal inference it is possible that I'm
    missing something.
5.  Alert the audience to potential confounding from unmeasured genetic
    influences that affect both parental and offspring behavioral and
    cognitive/affective disorders (see Appendix 4).
6.  Move the warnings about potential problems from recall bias to the
    introduction. Such bias seriously compromises the study's
    conclusions (see Appendix 5). Note that epidemiologists and medical
    researchers do not rely on memories when assessing the etiology of
    disease because memories are known to systematically bias results.
    Sickness tends to distort (or improve) reflections, leading to
    confounding measurement error.

Suggestions:

1.  Please spell out the acronyms. Personally, I'd suggest writing out
    ELA as "Early Life Adversity" because this is more readable.

2.  Please spell check the study and use consistent capitalisation.
    There were errors and inconsistencies on many pages.

3.  Please consider further "Specific Points to Consider".

## 

## Specific points to consider

-   Abstract should define jargon terms "internalising" and
    "externalising."
-   $n = 26,728$ should be $N = 26,728$
-   p.1, opening paragraph. Citation for claims about complex etiology
    of psychopathology should be included there.
-   p.3, middle of the page "... and evidence of a socialisation effect
    was demonstrated" Is "socialisation"the same as "externalisation?"
    Without definitions for key technical terms, your audience will be
    lost.
-   p.3 MDD -- acronym introduced on page 3 but it is not defined until
    p.8 (Major Depressive Disorder).
-   p.3 "...offspring were more likely to develop externalising symptoms
    when residing with the affected (AUD) parent" What does AUD mean?
    Acronym is never defined, or is buried. Again, suggest writing out
    all acronyms to ease your readers' cognitive burden.
-   p.3 "Additionally, having two affected parents substantially
    increased the risk of offspring developing externalising symptoms
    compared to offspring of one affected parent." What does
    "externalising symptoms"mean here? How are these attributes
    measured? Again, without the definitions your audience will be lost.
-   p.5 Are the terms "ELA"and "ACE" acronyms. If so, whey include both.
    If not, what are the differences? And where does "CM" fit in in?
    Again I find myself lost in the terminology.
-   p.18 "Vachon et al (2015a) note that few studies in this area have
    directly tested sex as a moderator, and those that have, report
    inconsistent results. Thus, our second aim is to add to the
    literature by including sex as a moderator in order to gain a
    clearer understanding of differential effects of parent and
    offspring sex." Note that stratifying on sex does enable causal
    inference about the role of effect modification by sex. And the
    correct way to test for moderation is to include the main effects
    and interactions of sex and ELA and the main effects and
    interactions of sex and parental internalisation/externalisation.
-   p.21 "Some data for our variables of interest were not available for
    unknown reasons." What effort did you make in requesting this
    information from NESARC? Detail about this effort needs to be
    included.
-   p.21 "In an effort to replicate our findings, as suggested by
    Pohlmann (2004), we randomly split our sample into two subsamples of
    13,364 participants each. These are referred to as Group One and
    Group Two. We then estimated the model twice, comparing the
    results." In cross validation, we split samples to *predict* outcome
    for the samples that were not used in estimation. I'd not heard of
    the mehod of splitting a sample and running a model twice so I
    tracked down the Pohlmann 2004
    [here](https://www.jstor.org/stable/pdf/27548054.pdf?refreqid=excelsior%3Ab06b63fd2f8576cc40592fc4834fb721).
    Pohlmann 2004 offers the following advice for exploratory factor
    analysis:

> Rather than using a priori sample-size guidelines, researchers should
> consider within-sample replication to gauge factor stability. This is
> accomplished by randomly dividing the sample into two halves and
> performing the same analysis on both halves. The researcher interprets
> the factor results that occur in both subsamples and ignores results
> that do not replicate. Result discrepancies are most likely caused by
> sampling error and should not be communicated to the readers.

Note that the recommended method is used to check the replicability of
data reduction, presumably in much smaller samples that what you have
here. Regarding the results presented here, differences in the results
of the two large "groups" suggest underlying problems with the models
and measures. I would simply run the model on the full sample and use
standard model fit evaluation, cross-validation, and robustness checks
to evaluate the results. What you've done here doesn't make a lot of
sense.

-   p 22. Missing data were excluded from the study, but on p.26 there
    is an account of how the missingness in NIAA used imputation for
    missingness in the NESARC-III. This is confusing.
-   p.15 "Data was gathered $\dots$" APA usage is "Data were gathered
    $\dots$"
-   p.25 "Variables were imputed using both assignment; in which
    relatable information is available on the same person record and
    analysts are confident about assigning values to items missing data,
    and allocation; in which the value for missing or inconsistent items
    cannot be derived from the same person record and is taken from
    other respondents with similar characteristics." Details of the
    (single) imputation method needed. More fundamental, if missing data
    were imputed how did you end up with missing data?
-   p\. 26 "the datum was" should be "the data were.''
-   p\. 26 "A detailed description of the imputation process and rates
    can be found in the National Epidemiologic Survey on Alcohol and
    Related Conditions-III (NESARC-III)- Data Notes (NIAAA,
    2014)."Please include a link.
-   p.27-28 "A new variable was computed for the final ELA category
    measured by yes/no responses by calculating the sum of variables.
    making up the category. The sum variable represents the number of
    different adverse environmental situations a participant had
    experienced, as opposed to intensity of the experienced adversity."
    This doesn't seem quite right. The responses here when summed are
    not counts or rates. They rather represent something like bandwidth
    for abuse. Are there any references you can use in support of the
    validity of indicator created through this method?
-   p.29. "To test measurement invariance we used the R function
    'measurementinvariance' which estimated and compared several
    versions of constaints." Which package. Also, what is "constaints"?
-   p.31. Structural equation model. I am curious how parental
    externalising/internalising could effect offspring
    externaling/internalising except through early life adversity?
    Indeed, Your results suggest this:

> While a significant direct effect did not exist between either
> Paternal Externalising and Externalising in females, or ELA and
> Externalising in females, results did indicate a significant total
> effect (standardised total coefficient .403, p=.004) of Paternal
> Externalising and ELA on Externalising in female offspring. p.38 mp.
> 38

-   p 36. Model fit CFI is .94, however, I'd thought your threshold was
    .95? (p.29). Ditto for page 37 results. Why is no explanation given?

-   Note, again, the following inference isn't correct. It belies a
    poor understanding of how mediation works. If there is full
    mediation the direct effects on the X$\to$Y path with drop out,
    while X$\to$M$\to$Y - indirect effect still holding.

>  We hypothesised Early Life Adversity would mediate the relationship
> between both Paternal and Maternal Externalising and latent
> Externalising and Internalising dimensions in offspring; that is ELA
> would account for the relationship between Parental Externalising, and
> Internalising and Externalising in male and female offspring. A
> significant indirect effect (standardised indirect coefficient .084,
> p=.001) was found between Paternal Externalising and Internalising
> mediated by ELA in male participants (Table 7). This result was
> replicated in Group Two males (see Appendix A, Table A12).\*\* Results
> also demonstrated a significant indirect effect (standardised indirect
> coefficient .103, p=.001) between Maternal Externalising and
> Internalising mediated by ELA in male offspring which was again
> replicated in Group Two males (Table 7, see Appendix A, Table A12).
> While a significant direct effect did not exist between either
> Paternal Externalising and Externalising in females, or ELA and
> Externalising in females, results did indicate a significant total
> effect (standardised total coefficient .403, p=.004) of Paternal
> Externalising and ELA on Externalising in female offspring. **This
> indicates that Paternal Externalising and ELA as a combination do have
> a significant association with Externalising in females.** (My
> emphasis.)


-   p.37 Perhaps I am missing something but I do not see differences in the Robust Comparative Fit Index (CFI), the Rbust Tucker-Lewis Index (TLI) or the Robust RMSEA for the constrained and  unconstrained models.

-  p.40 "No significant association between either Maternal or  Paternal Externalising and Internalising or Externalising in offspring was found in sons or daughters. As such, results do not provide support for our prediction." Again, if you remove ELA and Paternal Externalising and Internalising pop then this would suggest that Paternal effects are fully mediated by ELA. As far as I can tell from these results, you haven't actually tested the causal implications of the model you wish to test.


-   p.47 "In this study, parental externalising was operationalised through offspring reporting symptoms: participants were given a list of behaviours/situations that would indicate antisocial behaviour, or problematic substance use and were asked whether their parents engaged in those problematic behaviours. However, an affirmative response to parental externalising did not provide us with any information regarding the intensity or seriousness of the parental externalising." Again, the study's reliance on memory opens the door to confounding.

-   p.48 "Continuing this pattern of findings, our results do not demonstrate a significant association between externalising in either parent and internalising in offspring" -- however, this is *after* stratifying on ELA. And this makes perfect sense because what effect could parents have on offspring here except through ELA. Not genetic effects woudl act on parents and offspring, leading to confounding (see Appendix 4). 

- The point here seems to be underrated:

> This is a common issue faced in this area of research as genetic vulnerabilities and family adversity often overlap, particularly in early life, to increase risk; a phenomenon which has been termed the passive gene-environment correlation (Yan et al., 2020). By including such information in future analysis we would be able to infer not just the strength of the relationship between variables, but also more regarding the mechanisms by which they interact.

If genetic variation were to cause internalisation / externalisation in both parents and in offspring such effects would confounds inference. An instrumental variable would be needed to deconfound genetic influences (See Appendix 4).

p.59 It is true that sibling studies would be helpful. In addition to the points that you describes, sibling data would allow researchers to assess recall bias that is correlated with the outcomes variables.

-   p.61 Final sentence:

> Finally, though previous research has indicated that parental
> externalising has lasting adverse effects on offspring across their
> life-course, we found that these effects can be accounted for by
> experiences of ELA. In short, experiences of ELA, rather than parental
> externalising, appear to be the driving force behind the development
> of internalising symptoms in offspring.

However you has conditioned on a post-treatment variable. This isn't correct.

## Appendix 1: Simulation to demonstrate the post-treatment confounding

If the path from parental externalising/internalising (X) to offspring
externalising/internalising (Y) were fully mediated by early life
adversity (M), then including early life adversity in a statistical
model would cause the estimate from parent effects to offspring outcomes
to shrink towards zero -- as observed in Emerald's sudy. 

We would then infer that parental pathology causes offspring pathology
*because* parental pathology causes early life adversity.

To see this, let's simulate some data for a fully mediated relationship:

```{r}
# M -> Y 
# X -> M
set.seed(123)
n <- 1000
X <- rnorm( n )
M <- rnorm( n , X )
Y <- rnorm( n , M )

d_sim <- data.frame(X=X,M=M,Y=Y)
```

First, we condition only on X (parental effects), and find an effect of
X on Y (offspring internalisation/externalisation):

```{r}
ms1 <- parameters::model_parameters(
  lm(Y  ~ X, data = d_sim)
)
plot(ms1)
ms1
```

Next, assess only the mediator, early life adversity, and we find a
reliable association

```{r}
ms2<- parameters::model_parameters(
  lm(Y  ~ M, data = d_sim)
)
plot(ms2)
```

However, when we add both X and M, we find that X is attenuated.

```{r}
ms3<- parameters::model_parameters(
  lm(Y  ~ X + M, data = d_sim)
)
plot(ms3)
```

It seems as though parental influence has nothing to do with offspring
outcomes. However, we know that parental influence causes offspring
outcome because we've simulated the data to be that way.

Note that it doesn't help to run a 'mediation' model.

Mediation model

```{r}
library("brms")
library("rstan")
library("rstanarm")
library("tidybayes")
library("bayesplot")
library("easystats")
# rstan options
rstan_options(auto_write=TRUE)
options(mc.cores=parallel::detectCores ())
theme_set(theme_classic())


path_m <- brms::bf(M ~ X)
path_y <- brms::bf(Y ~ X + M)

m1 <- brms::brm(
  path_m + path_y + set_rescor(FALSE),
  data = d_sim,
  file = here::here("models","mediation-emerald")
)

parameters::model_parameters(m1)
```

Recalling:

```{r}
bmlm::mlm_path_plot(xlab = "Parental Causes\n(X)",
              mlab = "ELA\n(M)",
              ylab = "Offspring Effects\n(Y)")
```

We recover the indirect, direct and total effects as follows:

```{r}

post <- brms::posterior_samples(m1)
post2 <- post %>% 
  transmute(
    a = b_M_X  ,
    b = b_Y_M,
    cp =  b_Y_X,
    me = a * b,
    c = cp + me #,
  #  pme = me / c
  )
mcmc_intervals(post2)
```

Causal inference in observational studies is tricky. The key lesson here is that we should not condition on a post-treatment variable.

```{r}
# M -> Y 
# X -> M
set.seed(123)
n <- 1000
X <- rnorm( n )
M <- rnorm( n , X )
sex <- rep( 0:1 , each = n/2 )
Y <- rnorm( n , M ) * sex +  rnorm( n , M ) +  rnorm(n, sex)

d_sim2 <- data.frame(X=X,M=M,Y=Y,sex)

## Model
ms4<- parameters::model_parameters(
  lm(Y  ~ X + M*sex, data = d_sim2)
)
ms4
plot(ms4) + ggtitle("Simulated model including sex differences")
```

## Appendix 2: causal graph for the proposed model

The following is a directed acyclical graph (DAG) for the proposed model

```{r  layout="l-body-outset",  fig.height=8, fig.width=12}
# Graph model
library(ggdag)
dg_1 <- ggdag::dagify(
  outcome ~   sex + ela + mat_ex + mat_in + pat_ex + pat_in,
  ela ~   sex + mat_ex + mat_in + pat_ex + pat_in,
  exposure = "pat_ex",
  outcome = "outcome",
  labels = c(
    "outcome" = "outcome",
    "sex" = "sex",
    "ela" = "Early Life Adversity",
    "mat_ex" = "maternal external",
    "mat_in" = "maternal internal",
    "pat_ex" = "paternal external",
    "pat_in" = "paternal internal"
  ) )
# ) %>%
#   control_for("sex")

p1 <- ggdag(dg_1, text = FALSE, use_labels = "label") + labs(title = "Proposed Model") + theme_dag_blank()
p1
```

To obtain a valid causal inference we must:

1.  close all \`back doors'

2.  not introduce any collider confounds [@rohrer2018, @mcelreath2020,
    @pearl2019]

3.  not block any paths ELA and the Outcome
    (internalisation/externalisation)

To assess the covariates required for unbiased inference (conditional on
the assumed model), we evaluate that DAG:

```{r adjustsetDirect, layout="l-body-outset",  fig.height=8, fig.width=12}
# find adjustment set
ggdag::ggdag_adjustment_set(
  dg_1,
  exposure = "ela",
  outcome = "outcome",
  text = FALSE,
  use_labels  = "label"
) +
  theme_dag_blank() +
  labs(title = "Adjustment set",
  subtitle = "")
```

This is good news. All the variables in the assumed model are required
for robust estimation, and no new variables are required. Note that we
would need stratify on sex even if this were not the point of the study.

## Appendix 3: Casual inference where ELA fully mediates the causal effects of parent pathology on offspring pathology

Let's assume that the only way that parental internalisation or
externalisation can affect offspring internalisation or externalisation
is through early life adversity. Such a model appears to be consistent
with the results.

The assumed model is now:

```{r}
library(ggdag)
dg_2 <- ggdag::dagify(
  outcome ~   sex + ela,
  ela ~   sex + mat_ex + mat_in + pat_ex + pat_in,
  exposure = "ela",
  outcome = "outcome",
  labels = c(
    "outcome" = "outcome",
    "sex" = "sex",
    "ela" = "Early Life Adversity",
    "mat_ex" = "maternal external",
    "mat_in" = "maternal internal",
    "pat_ex" = "paternal external",
    "pat_in" = "paternal internal"
  )
) 

p2 <- ggdag(dg_2, text = FALSE, use_labels = "label") + labs(title = "Proposed Model") + theme_dag_blank()
p2
```

And the minimal adjustment set is now `sex`, which we must condition on
to assess the effect of ELA on offspring externalisation or
internalisation.

In this case we cannot obtain an estimate of parental influence on adult
offspring outcomes without introducing collider confounding.

```{r}
ggdag::ggdag_collider(
  dg_2,
  text = FALSE,
  use_labels  = "label"
) +
  theme_dag_blank() +
  labs(title = "Colliders bias is unavoidable in the fully mediated model",
  subtitle = "")
```

## Appendix 4: Causal inference and genetic confounding

Let's assume genetic influences on both parent and offspring
internalising/externalising. We use "G" to denote this influence:

```{r}
dg_4<- ggdag::dagify(
  outcome ~   sex + ela + G,
  ela ~   sex + mat_ex + mat_in + pat_ex + pat_in + G,
  mat_ex ~ G,
  pat_ex ~ G, 
  mat_in ~ G,
  mat_ex ~ G,
  exposure = "ela",
  outcome = "outcome",
  latent = "G",
  labels = c(
    "outcome" = "outcome",
    "sex" = "sex",
    "ela" = "Early Life Adversity",
    "mat_ex" = "maternal external",
    "mat_in" = "maternal internal",
    "pat_ex" = "paternal external",
    "pat_in" = "paternal internal",
    "G" = "(unmeasured) genetic influence"
  )
) 


ggdag(dg_4, text = FALSE, use_labels = "label") + labs(title = "ELA on outcome with latent genetic effects") + theme_dag_blank()


```

In this case, there would be no way to estimate the effects of parental
pathology on adult offspring pathology, regardless of mediation.

```{r}
ggdag::ggdag_adjustment_set(
  dg_4,
  text = FALSE,
  use_labels  = "label"
) +
  theme_dag_blank() +
  labs(title = "Adjustment set: ELA on outcome assuming genetic influences",
  subtitle = "")
```

## Appendix 5: The specter of measurement error

Finally, lets ignore genetic influence but assume measurement error. In
this case, we cannot remove the confounding influences of memory.

```{r}
dg_5<- ggdag::dagify(
  outcome ~   sex + ela,
  ela  ~   recollection + mat_ex + pat_ex + mat_in + pat_in, 
  recollection ~ outcome,
  mat_ex ~ recollection,
  pat_ex ~ recollection, 
  mat_in ~ recollection,
  pat_in ~ recollection, 
  exposure = "ela",
  outcome = "outcome",
  latent = "recollection",
  labels = c(
    "outcome" = "outcome",
    "sex" = "sex",
    "ela" = "Early Life Adversity",
    "mat_ex" = "maternal external",
    "mat_in" = "maternal internal",
    "pat_ex" = "paternal external",
    "pat_in" = "paternal internal",
    "recollection" = "recollection  bias"
  )
) 
ggdag::ggdag_adjustment_set(
  dg_5,
  text = FALSE,
  use_labels  = "label"
) +
  theme_dag_blank() +
  labs(title = "Adjustment set: assume recollection bias",
  subtitle = "")
```

## Packages used

```{r echo = FALSE}
report::cite_packages()
```
