---
title: "Review of Laural Keat's Memory Study"
description: |
author:
  - name: Joseph Bulbulia
    url: https://josephbulbulia.netlify.app
    affiliation: Victoria University of Wellington
    affiliation_url: https://www.wgtn.ac.nz
    orcid_id: 0000-0002-5861-2056
date: June 2, 2021
output:
  distill::distill_article:
    self_contained: true
    toc: false
    code_folding: true
    highlight: tango
    highlight_downlit: false
bibliography: references.bib
---

```{r setup, include = FALSE}
# set up
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = TRUE
)
```

```{r message = FALSE}
# required packages
if (!require(ggdag)) install.packages("ggdag")
```

## Causal graphs

Externalising

```{r graph1, layout="l-body-outset",  fig.height=8, fig.width=12}
# Graph model
library(ggdag)
dg_1<- ggdag::dagify(
  Externalising ~   Early_Life_Adv  + Father_Externalising + Father_Internalising + Mother_Externalising + Mother_Internalising, 
  Early_Life_Adv ~  Sex + Father_Externalising + Father_Internalising + Mother_Externalising + Mother_Internalising, 
  exposure = "Early_Life_Adv",
  outcome = "Externalising",
  labels = c(
    "Externalising" = "Externalising",
    "Early_Life_Adv" = "Early_Life_Adversity",
    "Father_Externalising" = "Father_Externalising",
    "Father_Internalising" = "Father_Internalising",
    "Mother_Externalising" = "Mother_Externalising",
    "Mother_Internalising" = "Mother_Internalising",
    "Sex" = "Sex"
   )
)%>%
  control_for(c("Sex"))

p1 <-ggdag(dg_1, text = FALSE, use_labels = "label") + theme_dag_blank() + labs(title = "Proposed Causal Model for Externalising")
p1
```

Internalising:

```{r  graph2, layout="l-body-outset",  fig.height=8, fig.width=12}
# Graph model
library("ggdag")

dg_2<- ggdag::dagify(
  Internalising ~   Early_Life_Adv  + 
    Father_Externalising + 
    Father_Internalising + 
    Mother_Externalising +
    Mother_Internalising, 
  Early_Life_Adv ~  Sex + 
    Father_Externalising + 
    Father_Internalising + 
    Mother_Externalising + 
    Mother_Internalising, 
  exposure = "Early_Life_Adv",
  outcome = "Internalising",
  labels = c(
    "Internalising" = "Internalising",
    "Early_Life_Adv" = "Early_Life_Adversity",
    "Father_Externalising" = "Father_Externalising",
    "Father_Internalising" = "Father_Internalising",
    "Mother_Externalising" = "Mother_Externalising",
    "Mother_Internalising" = "Mother_Internalising",
    "Sex" = "Sex"
   )
)%>%
  control_for("Sex")

p2 <-ggdag(dg_2, text = FALSE, use_labels = "label") + theme_dag_blank() + labs(title = "Proposed Causal Model for Internalising")
p2
```

\
\

## Test for Externalising

1.  It is important to close all \`back doors'

2.  It is *equally* important introduce any collider confounds open. This latter point is under-appreciated, however see: [@rohrer2018, @mcelreath2020, @pearl2019]. Note that [@rohrer2018] is very accessible.

3.  We should not block any paths between episodic memory at time 1 and depression at time 2.

The following are the potential adjustment sets that enable unbiased inference conditional on our model assumptions:

```{r adjustsetDirect, layout="l-body-outset",  fig.height=8, fig.width=12}
# find adjustment set
p2 <-ggdag::ggdag_adjustment_set(
  dg_1,
  exposure = "Early_Life_Adv",
  outcome = "Externalising",
  text = FALSE,
  use_labels  = "label"
) +
  theme_dag_blank() +
  labs(title = "Adjustment set",
  subtite = "")
p2
```

## Test for Internalising

Likewise, to test the causal effect of Early Life Adversity while stratifying on Sex, it is correct to adjust for Father Externalising and Mother Externalising

`{# find adjustment set} p3 <-ggdag::ggdag_adjustment_set(   dg_2,   exposure = "Early_Life_Adv",   outcome = "Internalising",   text = FALSE,   use_labels  = "label" ) +   theme_dag_blank() +   labs(title = "Adjustment set",   subtite = "") p3`

## Lurking Problem

## NOTES

```{r graph1, layout="l-body-outset",  fig.height=8, fig.width=12}
# Graph model
library(ggdag)
dg_1<- ggdag::dagify(
  Depression2 ~   Memory1  + Depression1, 
  Memory2 ~ Memory1 + Depression2,
  exposure = "Memory1",
  outcome = "Depression2",
  labels = c(
    "Depression2" = "Depression2",
    "Memory1" = "Memory1",
    "Memory2" = "Memory2",
    "Depression1" = "Depression1"
   )
)#%>%
  #control_for(c("Memory2"))

p1 <-ggdag(dg_1, text = FALSE, use_labels = "label") + theme_dag_blank() + labs(title = "Proposed Causal Model for Externalising")
p1
ggdag::ggdag_collider(dg_1) + theme_dag_gray()

ggdag::ggdag_adjustment_set(dg_1) + theme_dag_gray()

```

## Packages used

```{r echo = FALSE}
report::cite_packages()
```
